---
title: "ArHa manuscript 1: A global synthesis of rodent- and shrew-borne Arenaviruses and Hantaviruses"
format: html
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)

```

```{r}
#| label: abstract
#| include: false

data <- read_rds(here("data", "database", "Project_ArHa_database_2025-08-20.rds"))

included_studies <- data$descriptives
n_included_studies <- included_studies |>
  distinct(full_text_id) |>
  nrow()

citations <- data$citations |>
  filter(full_text_id %in% included_studies$full_text_id)

min(citations$publication_year)
max(citations$publication_year)

host <- data$host |>
  ungroup()
n_rodents_sampled <- host |>
  filter(str_detect(host_order, "Rodentia")) |>
  summarise(n_detected = sum(number_of_hosts, na.rm = TRUE))
n_shrews_sampled <- host |>
  filter(str_detect(host_order, "Soricomorpha")) |>
  summarise(n_detected = sum(number_of_hosts, na.rm = TRUE))
n_others_sampled <- host |>
  filter(!str_detect(host_order, "Rodentia") &
    !str_detect(host_order, "Soricomorpha")) |>
  summarise(n_detected = sum(number_of_hosts, na.rm = TRUE))

pathogen <- data$pathogen 
n_assayed <- pathogen |>
  summarise(number_tested = sum(number_tested, na.rm = TRUE))
n_pathogen_records <- pathogen |>
  nrow()

sequence <- data$sequence |>
  distinct(accession_primary, .keep_all = TRUE)
table(sequence$sequence_type)
```



# Abstract

Rodents and shrews are key reservoirs for several zoonotic viruses in the families *Arenaviridae* and *Hantaviridae*, including pathogens causing viral haemorrhagic fevers with significant public health impact. Yet, our understanding of host-virus associations, discovery effort, and global sampling gaps remains fragmented. We conducted a systematic review of Arenavirus and Hantavirus detection in small mammals from 1977 to 2025, identifying 753 studies reporting on over 579,000 rodents (Order: Rodentia), 12,100 shrews (Order: Eulipotyphla; formerly Soricomorpha) and 2,500 other mammals. 

From these, we extracted over 715,000 assays comprising 69,993 pathogen detection records, harmonised to the highest spatiotemporal and taxonomic resolution. We further linked 5,651 pathogen RNA sequences and 2,406 host DNA sequences from GenBank with associated assay and sampling metadata. The result is a global, open-access dataset capturing small mammal testing effort, pathogen detection and sequence availability.

We describe host-virus associations and use co-phylogenetic analysis to examine evolutionary congruence between pathogens and hosts. We assess the geographic and taxonomic sampling biases and identify undersampled regions and clades with high potential for viral discovery. Comparative analyses of virus with and without known zoonotic potential highlight traits associated with host breadth and spillover risk. This harmonised resource enables new insights into the ecology and evolution of rodent- and shrew-borne viruses and offers a foundational resource for prioritising surveillance and forecasting zoonotic emergence.

# Introduction

+ Broad Context: Arenaviruses and hantaviruses, hosted by small mammals (Rodentia and Eulipotyphla), are globally distributed and include significant zoonotic pathogens.
The Classic View vs. New Evidence: Historically, these viruses were considered to have tight, co-evolved relationships with single host species. However, increasing surveillance efforts reveal a more complex picture, with a growing number of host species associated with each pathogen.
+ The Central Problem - The "Surveillance Landscape": Our understanding of host range, viral sharing, and zoonotic risk is fundamentally shaped by a biased "surveillance landscape." The true ecological picture is obscured by where, when, and for what we choose to look. An understanding of the host range of each viral species is driven by research effort, or more accurately, sampling effort across the small-mammal community in the endemic region of the pathogen.
+ The Knowledge Gap: The role of these multiple potential host species in sustaining transmission and their capacity for zoonotic spillover remains an open and critical question. We lack a quantitative framework to map the boundaries of our collective knowledge and identify the key taxonomic and geographic gaps in global surveillance.
+ Study Aim: This study addresses this gap by synthesizing the global literature on Arenaviridae and Hantaviridae in wild small mammals. We construct a comprehensive, spatio-temporally explicit database to:
    + Describe the current state of global surveillance data.
    + Quantify the taxonomic and geographic biases inherent in this data.
    + Provide an evidence-based roadmap for future research and pandemic preparedness by identifying critical surveillance gaps.
    + Use this synthesized data to test initial hypotheses about the ecological drivers of host-pathogen associations.

# Methods

##  Systematic review

To synthesise available data on the sampling of Arenviridae and Hantaviridae in wild-living animals we conducted a systematic review of the literature.

We searched NCBI PubMed and Clarivate Web of Science using the following terms.

+ Rodent* OR Shrew
+ Arenvir* OR Hantavir*
+ a AND b

Searches were conducted on YYYY-MM-DD

Inclusion and exclusion criteria were as follows:

+ Inclusion
+ Exclusion

A total of A citations were returned; after de-duplication, B citations were screened, supplemented with C manuscripts from citation chaining. Titles and abstracts were screened, with D records carried forward to full-text review. Of these, E manuscripts contained data suitable for extraction.

##  Dataset construction

We developed a relational data extraction tool to produce four linked datasets: Descriptive, Host, Pathogen, and Sequence.

Data was aligned with Darwin Core Terms to facilitate future deposition to the Global Biodiversity Information Facility (GBIF).

Key extracted variables included: publication metadata, sampling effort, host species and abundance, sampling location and time, pathogen assay results (tested, positive), and GenBank accession numbers.

##  Analysis of Taxonomic and Spatio-Temporal Biases

### Host sampling biases

+ We summarised sampling events, individuals sampled, and pathogen detections across host taxa (genus, species).
+ To quantify taxonomic bias, we compared the observed sampling distribution against a null model of even sampling across all known species within Rodentia and Eulipotyphla.
+ We fitted generalized additive models (GAMs) to test whether host traits (e.g., body mass, geographic range size, biome ranges, synanthropy, IUCN threat status) were associated with sampling intensity.

### Pathogen Sampling Biases

+ We examined the diversity of Arenaviridae and Hantaviridae viruses detected and compared the frequency of detection for each virus species against the total sampling effort in its known host taxa.
+ We quantified the frequency of co-detection, recording how often multiple virus species were screened for within a single study or host population.
+ For virus-positive samples with available sequences, we assessed the representativeness of genetic meta-data in relation to the geographic and host sampling coverage.

### Geographic Sampling Biases

+ We mapped sampling locations against the known geographic ranges of key host genera (e.g., Apodemus, Mastomys) from the IUCN Red List to visually and quantitatively identify areas of host presence with no reported sampling effort.
+ We correlated country-level sampling intensity (number of studies and animals sampled) with socio-economic indicators (e.g., GDP per capita, national research funding) to test the hypothesis that surveillance effort is driven by resource availability.

### Temporal Sampling Biases

+ We quantified sampling effort (number of studies, hosts sampled) over time, aggregated by year and decade of sample collection.
+ We analyzed trends in the diversity of hosts and pathogens sampled over time.
+ We qualitatively assessed whether major historical events (e.g., the 1993 Sin Nombre virus outbreak, the development of new diagnostic techniques) corresponded with subsequent increases in sampling effort or the diversity of detected viruses.

##  Bipartite network analysis of host–pathogen associations

### Descriptive Network Analysis

 + We constructed bipartite (host–virus) networks separately for Arenaviridae and Hantaviridae. To account for diagnostic uncertainty, these networks were stratified by detection method:
    + "Acute Evidence" Network: Including only direct detection methods (PCR, culture, sequencing).
    + "All Evidence" Network: Including both direct and indirect detection methods (serology).
 + In these networks, nodes represent host species and virus species, and edges represent reported detections.
 + We calculated key network-level metrics (e.g., connectance, modularity, nestedness) for both virus families and formally compared their structures across the two evidence strata.
 + We calculated node-level metrics (degree) to identify potential "keystone" taxa, such as highly connected generalist hosts or hub pathogens.

### Hypothesis Testing with Exponential Random Graph Models (ERGMs)

 + To move beyond description and formally test hypotheses about the drivers of host-pathogen interactions, we employed an ERGM framework. This approach allows us to model the probability of a host-pathogen link while explicitly controlling for sampling effort.
 + Network Construction for ERGMs: We constructed a comprehensive host x pathogen incidence matrix for each virus family. For each host-pathogen dyad, the value was coded as 1 if the host was tested and found positive, 0 if the host was tested and found negative (a true non-detection), and NA if the host was never tested for that pathogen. This incorporation of non-detection data is critical for an unbiased analysis.
 + Model Formulation: Using the statnet suite of packages in R, we fitted ERGMs to model the probability of a tie (an infection). The models included terms to test our key hypotheses:
    + Sampling Effort: We included the total number of individuals tested for each host species as a node-level covariate (nodecov) to control for sampling bias.
    + Ecological Traits: We tested the significance of host traits as predictors of infection. Key covariates included geographic range size and a categorical variable for synanthropy (commensal vs. wild species).|
 +Objective: The primary goal of the ERGM analysis was to determine if ecological traits of a host species significantly predict its likelihood of being a host for a given virus, after accounting for the fact that some species are much more heavily sampled than others.

# Results

##  Description of the Dataset

+ Summary of the total number of unique studies, sampling sites, and years represented.
+ Overview of the number of host species sampled and their distribution across families and genera.
+ Summary of the number and diversity of viruses tested, detected, and sequenced.

##  The Surveillance Landscape: Quantified Biases

+ Taxonomic Gaps: Presentation of results showing over- and under-sampling of specific host and pathogen taxa. Identification of key host genera and virus species that are under-represented relative to their known diversity or host associations.
+ Geographic Gaps: Presentation of maps and statistical results highlighting major geographic regions where known hosts exist but sampling is absent. Results of the correlation between sampling effort and socio-economic factors.
+ Temporal Trends: Presentation of plots showing trends in sampling effort and detected diversity over time, highlighting periods of intensified research.

##  Host-Pathogen Network Structures

+ Visualization and description of the Arenaviridae and Hantaviridae networks.
+ Results of the quantitative comparison of the two network structures.
+ Identification of keystone host and pathogen taxa.

# Discussion

Our synthesis reveals a global surveillance landscape for rodent-borne viruses that is highly fragmented and biased across taxonomic, geographic, and temporal dimensions. Surveillance is concentrated in specific taxa, regions, and time periods, leaving vast "known unknowns."

An Evidence-Based Roadmap for Future Surveillance:

+ Synthesize the taxonomic, geographic, and temporal gap analyses to provide a clear, actionable "roadmap."
+ Explicitly list the top host genera, geographic regions, and under-studied time periods that represent the most critical priorities for future surveillance efforts.

Ecological Insights from Network Analysis: Discuss the implications of the observed network structures. For example, do the differences between the Arena- and Hantavirus networks suggest different evolutionary strategies or transmission dynamics?

Limitations: Acknowledge limitations of the systematic review process (e.g., publication bias, language limitations, inability to capture all "grey literature").

Conclusion: This work provides both a valuable, novel data resource and a quantitative framework for understanding the landscape of zoonotic virus surveillance. By identifying where, in which species, and when we are not looking, we provide a crucial tool for guiding more strategic and efficient global health research and pandemic preparedness.
