---
title: "Study protocol: Production of Arenavirus and Hantavirus host-pathogen database."
author:
  - name: "David Simons"
    url: https://www.dsimons.org
    affiliation: The Royal Veterinary College, London, UK
    affiliation_url: https://www.rvc.ac.uk
    orcid_id: 0000-0001-9655-1656
date: "`r Sys.Date()`"
categories:
  - Arenavirus
  - Hantavirus
---

``` {r setup, include = FALSE}

if (!require("pacman")) install.packages("pacman")
pkgs =
  c("here",
    "tidyverse",
    "DT"
  )

pacman::p_load(pkgs, character.only = T)

```


# Motivation

# Introduction

# Project aims

# Method

## Search strategy

An initial search was run on NCBI Pubmed 2023-01-06.

  1. rodent* - 165,513
  2. arenavir* - 2,367
  3. hantavir*.mp. - 4,022
  4. 2 OR 3 - 6,308
  5. 1 AND 4 - 1,842

Citations were downloaded as a text file and imported into `R` for processing. Deduplication by Pubmed ID resulted in 1821 distinct citations.

``` {r search-results}

search_date <- "2023-01-06"

articles <- read_rds(here("data", "search", paste0("citations_", search_date, ".rds")))

articles

```

An initial search of the returned citations was conducted to identify ten studies to trial data extraction. 

## Data extraction

## Data formatting