---
title: "Study protocol: Production of Arenavirus and Hantavirus host-pathogen database."
author:
  - name: "David Simons"
    url: https://www.dsimons.org
    affiliation: The Royal Veterinary College, London, UK
    affiliation_url: https://www.rvc.ac.uk
    orcid_id: 0000-0001-9655-1656
date: "`r Sys.Date()`"
format: 
  html: 
    toc: true
    toc-location: left
    code-fold: true
    code-summary: "Code"
    page-layout: full
crossref:
  fig-title: '**Figure**'
categories:
  - Arenavirus
  - Hantavirus
---

``` {r setup, include = FALSE}

if (!require("pacman")) install.packages("pacman")

pkgs =
  c("caret",
    "colorspace",
    "here",
    "htmltools",
    "leaflet",
    "plotly",
    "sf",
    "tidyverse"
  )

pacman::p_load(pkgs, character.only = T)

```


# Motivation

Current host-pathogen association datasets provide synthesised information on hosts and their pathogens but do not contain temporal or geographic information. These resources often provide linking information to publications reporting the association but information including accession numbers of archived sequences, number of individuals tested and measures of prevalence among sampled populations are not immediately retrievable. Using these resources for inference beyond host-pathogen associations is therefore limited. Here, we aim to produce a database of host-pathogen associations for two viral families of small mammals which contain several known zoonoses, namely Arenaviridae and Hantaviridae. This database can be used to explore the distribution of small mammal hosts of known and suspected pathogens and the extent to which they have been sampled. Further, linkage to sequence data of known zoonoses will support analysis of the risk of viral reassortment between viral species within geographically co-located host species.

# Introduction



# Project aims

  1. Review the literature to produce a synthesised dataset on Arenaviridae and Hantaviridae among small mammals
  2. Implement visualisation tools to explore sampling extent for hosts and pathogens to further understanding of host-pathogen associations
  3. Investigate the relative contributions of ecological, geographic, and genomic factors leading to cross-species transmission and reassortment in rodent-associated arenaviruses and hantaviruses

# Method

## Search strategy

An initial search was run on NCBI Pubmed 2023-01-06.

  1. rodent* - 165,513
  2. arenavir* - 2,367
  3. hantavir*.mp. - 4,022
  4. 2 OR 3 - 6,308
  5. 1 AND 4 - 1,842

Citations were downloaded as a text file and imported into `R` for processing. Deduplication by Pubmed ID resulted in 1821 distinct citations.

``` {r search-results}

search_date <- "2023-01-06"

articles <- read_rds(here("data", "search", paste0("citations_", search_date, ".rds")))

articles

```

An initial search of the returned citations was conducted to identify nine studies to trial data extraction. 

### Inclusion criteria

Studies were included if they contained information on:

  1. Rodent sampling, through direct or indirect detection of microorganisms in the Arenaviridae and Hantaviridae families AND
  2. Information on the trapping location of rodents sampled

Studies were excluded if they did not contain information on the rodent source of the sample for direct or indirect detection of Arenaviridae or Hantaviridae. Studies reporting experimental infections and human infections were also excluded.
  
Direct detection was defined as PCR or culture of virus and indirect detection was defined as detection of antibodies against a Arenaviridae or Hantaviridae.

## Data extraction

### Included studies

Information about the included study is extracted in a `descriptive` sheet.

| Column name | Description |
| ---- | :-------- |
| study_id | A unique identifier for the included study |
| pubmed_id | The pubmed ID of the included study (if available) |
| DOI | The digital object identifier of the study (if available), if a DOI is not available a weblink to a persistent page for the reference will be included |
| first_author_surname | The first authors surname |
| title | The title of the manuscript, report or book section |
| journal | The name of the journal, report or book |
| year | The year of publication |
| study_design | A free text entry succinctly describing the study design for the rodents or pathogens |
| sampling_effort | A free text entry to capture the effort of sampling, ideally in trap nights for rodent studies |
| data_access | Whether study data is available in complete form or whether summarise data only are available |
| linked_manuscripts | The DOI or weblink to other studies including the same dataset either in its entirety or a subset, this will be used to attempt to de-duplicate data |

: Study information extraction sheet {#tbl-description}

### Rodent sampling

Rodent data is extracted in a `rodent` sheet. This will include information on the timing of data collection, the location of data collected, the small mammal species detected either as detection/non-detection or number detected. For studies not reporting the detection of an individual species at a location that has been detected at other locations the species will be entered as not-detected at that location.

| Column name | Description |
| ---- | :-------- |
| rodent_record_id | A unique identifier for the rodent species, at a specific location or timepoint reported by the study dependent on the level of aggregation reported in the study |
| study_id | A unique identifier to link a study to the `descriptive` sheet entry for that study |
| date | The period in which data collection of rodent samples occurred, this will be extracted at the highest temporal resolution provided |
| genus | The genus of the small-mammal as reported in the study |
| species | The species of the small-mammal as reported in the study |
| location | The location of sampling effort, depending on how data are presented in the study this will match to the coordinates given for trapping effort. i.e. if trapping is aggregated at village level, village names will be used |
| country | Country where trapping occurred, for multinational studies where numbers are not disaggregated by country, all countries will be included |
| habitat_type | High level habitat type will be recorded here at the scale for which trapping is recorded |
| coordinate_resolution | The description of coordinate levels provided in the study, i.e. aggregated at study site or study region |
| latitude | Latitude will be converted from coordinates presented to EPSG:4326 |
| longitude | Longitude will be converted from coordinates presented to EPSG:4326 |
| number | The number of detected individuals, for capture-mark-recapture studies the number of distinct individuals will be entered. For studies not explicitly reporting non-detection, values of 0 for a species or genus will be entered if it is detected elsewhere in the study |

: Rodent information extraction sheet {#tbl-rodent}

### Pathogen sampling

Pathogen assays are extracted in the `pathogen` sheet. This includes information on the host the sample originated from, the pathogen family and species being assayed for and the method of the assay. For studies conducting multiple assays on the same samples for different pathogens additional records will be added for each assay. Similarly, if antibody and direct detection for the same pathogen on the same samples is performed additional records will be added. 

| Column name | Description |
| ---- | :-------- |
| pathogen_record_id | A unique identifier for the group of samples from the same rodent species, at a specific location or timepoint, tested for the same pathogen using the same method |
| study_id | A unique identifier to link a study to the `descriptive` sheet entry for that study |
| date | The period in which data collection of rodent samples occurred, this will be extracted at the highest temporal resolution provided |
| host_genus | The genus of the small-mammal from which the sample originated as reported in the study |
| host_species | The species of the small-mammal from which the sample originated as reported in the study |
| location | The location of sampling effort, depending on how data are presented in the study this will match to the coordinates given for trapping effort. i.e. if trapping is aggregated at village level, village names will be used |
| country | Country where trapping occurred, for multinational studies where numbers are not disaggregated by country, all countries will be included |
| habitat_type | High level habitat type will be recorded here at the scale for which trapping is recorded |
| coordinate_resolution | The description of coordinate levels provided in the study, i.e. aggregated at study site or study region |
| latitude | Latitude will be converted from coordinates presented to EPSG:4326 |
| longitude | Longitude will be converted from coordinates presented to EPSG:4326 |
| pathogen_family | The family of virus being assayed for (i.e. Arenaviridae or Hantaviridae) |
| pathogen_species | The species of virus being assayed for if a specific test is being used. For assays unable to differentiate between multiple viral species `Multiple` will be entered |
| detection_method | Whether the assay is attempting to detect antibody, direct detection of virus (i.e. pcr), or other |
| number_tested | The number of distinct samples tested |
| number_negative | The number of reported negative samples |
| number_positive | The number of reported positive samples |
| number_inconclusive | The number of samples with inconclusive results |

: Pathogen information extraction sheet {#tbl-pathogen}

### Pathogen sequences

If studies include linkage to complete or partial sequences of viruses archived in NCBI they will be lined through the `pathogen_sequences` sheet. 

| Column name | Description |
| ---- | :-------- |
| sequence_record_id | A unique identifier for the sequence record |
| study_id | A unique identifier to link a study to the `descriptive` sheet entry for that study |
| host_genus | The genus of the small-mammal from which the sample originated as reported in the study |
| host_species | The species of the small-mammal from which the sample originated as reported in the study |
| pathogen_species | The species of the pathogen |
| accession_number | The accession number for each record archived by the study |

: Pathogen sequences extraction sheet {#tbl-sequence}

### Zoonosis status

Finally, an additional sheet `known_zoonoses` will be produced containing all of the viral species sampled and whether they are known to cause disease in humans.

| Column name | Description |
| ---- | :-------- |
| pathogen_id | A unique identifier for the pathogen species |
| pathogen_family | The family of pathogen |
| pathogen_species | The species of pathogen |
| known_zoonosis | A logical statement of pathogenicity among humans |
| disease_name | The disease name caused by the viral species, if multiple diseases they are entered with a comma separator |
| icd_10 | The ICD-10 associated name of the disease (if known) |
| disease_reference | A DOI of a publication that supports the `known_zoonosis` statement |

: Zoonosis status extraction sheet {#tbl-sequence}

## Data processing

Raw data will be downloaded from Google Sheets using the `googledrive` API in `R`, with date stamped files stored locally. Data will be imported into `R` for processing, cleaning and formatting to produce a dataset suitable for further analysis.

# Outputs

## Data visualisation

An `RShiny` web-based application will be produced to visualise the database and support future analysis.

### Rodent sampling

An example of a map produce from rodent sampling data included in the initial 9 studies is shown below.

```{r leaflet-rodents, warning = FALSE}
#| fig-height: 8
#| label: rodent-map
#| fig-cap: An interactive map displaying the location of detected small species in included studies. Selecting points will expand the number of data points at those coordinates. Point colour indicates small mammal genus with size of the point varying by the number of individuals detected. Hovering over a point or selecting it will show the species name, the number detected and the time period surveyed at the location. Data is currently shown for 9 studies.

combined_data <- read_rds(here("data", "raw_data", paste0(search_date, "_data.rds")))

project_crs <- "EPSG:4326"

rodent_locations <- combined_data$rodent %>%
  select(genus, species, latitude, longitude, number) %>%
  mutate(number = as.numeric(number)) %>%
  drop_na(number) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = project_crs) %>%
  group_by(geometry, species) %>%
  mutate(number = sum(number)) %>%
  ungroup() %>%
  distinct() %>%
  rowwise() %>%
  mutate(labels = HTML(paste0("Species: ", species, "<br>", "Number detected: ", number)))

pal <- colorFactor(diverging_hcl(n = length(sort(unique(rodent_locations$genus))), palette = "Berlin"),
                   domain = sort(unique(rodent_locations$genus)))

leaflet(data = rodent_locations) %>%
  addTiles() %>%
  addCircleMarkers(radius = ~ifelse(sqrt(number) == 0, 1, sqrt(number)),
                   color = ~pal(genus),
                   stroke = FALSE,
                   fillOpacity = 0.8,
                   clusterOptions = markerClusterOptions(spiderfyOnMaxZoom = TRUE,
                                                         spiderLegPolylineOptions = list(length = 6),
                                                         spiderfyDistanceMultiplier = 2),
                   label = ~labels,
                   popup = ~labels) %>%
  addLegend("bottomright", pal = pal, values = ~genus, title = "Host genus", opacity = 0.8)

```


### Pathogen detection

A similar map can be produced to map pathogen detection with separate layers for acute infection (i.e. PCR positive samples) and evidence of prior infection (i.e. antibody positive samples)

``` {r leaflet-pathogens, warning = FALSE}
#| label: pathogen-map
#| fig-cap: An interactive map showing the locations of pathogen sampling. Selecting clustered points will expand the data. Each individual circle represents a single host species pathogen combination at that sampling location. The number of samples tested is associated with circle size. Selecting the point will display information about the host-pathogen association, including number of samples tested and number of positive samples. A white border of the circle designates host-pathogen associations with at least one positive result. The colour of the circle indicates the pathogen species (where known). Two layers are available, for viral detection and antibody detection.


pathogen_locations <- combined_data$pathogen %>%
  select(detection_method, pathogen_family, pathogen_species, host_species, latitude, longitude, number_tested, number_positive) %>%
  mutate(number_tested = as.numeric(number_tested),
         number_positive = as.numeric(number_positive)) %>%
  drop_na(number_tested, number_positive) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = project_crs) %>%
  group_by(geometry, detection_method, pathogen_family, pathogen_species, host_species) %>%
  mutate(number_tested = sum(number_tested),
         number_positive = sum(number_positive)) %>%
  ungroup() %>%
  distinct() %>%
  mutate(percent_positive = case_when(is.nan(number_positive/number_tested * 100) ~ "NA",
                                      number_positive == 0 ~ "NA",
                                      number_positive/number_tested * 100 < 1 ~ "< 1%",
                                      TRUE ~ paste0(as.character(round(number_positive/number_tested * 100, 2)), "%")),
         detection_method = case_when(detection_method == "pcr" ~ "Viral",
                                      detection_method == "antibody" ~ "Antibody"),
         detection = case_when(number_positive >= 1 ~ "Detected",
                               TRUE ~ "Not detected")) %>%
  rowwise() %>%
  mutate(labels = HTML(paste0("Pathogen family: ", pathogen_family, "<br>",
                              "Pathogen species: ", pathogen_species, "<br>",
                              "Host species: ", host_species, "<br>",
                              "Number tested: ", number_tested, "<br>",
                              "Number positive (%): ", number_positive, " (", percent_positive, ")", "<br>")))


path_pal <- colorFactor(diverging_hcl(n = length(unique(pathogen_locations$pathogen_species)), palette = "Berlin"),
                        domain = sort(unique(pathogen_locations$pathogen_species)))
stroke_pal <- colorFactor(palette = c("white", "gray"), levels = c("Detected", "Not detected"))

leaflet(data = pathogen_locations) %>%
  addTiles() %>%
  addCircleMarkers(data = pathogen_locations %>%
                     filter(detection_method == "Viral") %>%
                     filter(number_tested != 0),
                   radius = ~ifelse(sqrt(number_tested) == 0, 1, sqrt(number_tested)),
                   fill = TRUE,
                   fillColor = ~path_pal(pathogen_species),
                   stroke = TRUE,
                   color = ~stroke_pal(detection),
                   weight = 2,
                   opacity = 1,
                   fillOpacity = 0.8,
                   clusterOptions = markerClusterOptions(spiderfyOnMaxZoom = TRUE,
                                                         spiderLegPolylineOptions = list(length = 6),
                                                         spiderfyDistanceMultiplier = 2),
                   label = ~labels,
                   popup = ~labels,
                   group = "Viral detection") %>%
  addCircleMarkers(data = pathogen_locations %>%
                     filter(detection_method == "Antibody") %>%
                     filter(number_tested != 0),
                   radius = ~ifelse(sqrt(number_tested) == 0, 1, sqrt(number_tested)),
                   fill = TRUE,
                   fillColor = ~path_pal(pathogen_species),
                   stroke = TRUE,
                   color = ~stroke_pal(detection),
                   weight = 2,
                   opacity = 1,
                   fillOpacity = 0.8,
                   clusterOptions = markerClusterOptions(spiderfyOnMaxZoom = TRUE,
                                                         spiderLegPolylineOptions = list(length = 6),
                                                         spiderfyDistanceMultiplier = 2),
                   label = ~labels,
                   popup = ~labels,
                   group = "Antibody detection") %>%
  addLayersControl(overlayGroups = c("Viral detection", "Antibody detection"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  addLegend("bottomright", pal = path_pal, values = ~pathogen_species, title = "Pathogen species", opacity = 0.8)

```

### Host pathogen associations

``` {r  plotly-hp, warning = FALSE}
#| fig-width: 9
#| fig-height: 12
#| label: hp-associations
#| fig-cap: An interactive plot displaying observed host pathogen associations. Facets are produced for antibody based assays and viral detection assays. The rodent species are listed in alphabetical order on the y-axis, detected pathogens are displayed in alphabetical order on the x-axis. Purple tiles represent host-pathogen associations that were not observed for that detection type, yellow tiles represent observed host-pathogen associations. The strength of the colour (the alpha) is scaled to the number of assays performed for that host-pathogen pair. A black line surrounding the tile indicates the pathogen as a known zoonosis. Hovering over the relavent tile will highlight the number of assays and the number of positive samples.

pathogen_summary <- combined_data$pathogen %>%
  select(genus = host_genus,
         species = host_species,
         pathogen_family, pathogen_species,
         number_tested, number_positive,
         detection_method) %>%
  mutate(number_tested = as.numeric(number_tested)) %>%
  filter(number_tested >= 1) %>%
  left_join(combined_data$zoonoses %>%
              select(pathogen_species, known_zoonosis),
            by = c("pathogen_species"))

hp_associations <- combined_data$rodent %>%
  select(genus, species) %>%
  left_join(pathogen_summary %>%
              group_by(detection_method, species, pathogen_species, known_zoonosis) %>%
              mutate(number_tested = sum(number_tested),
                     number_positive = sum(number_positive)), by = c("genus", "species")) %>%
  drop_na(number_tested) %>%
  distinct() %>%
  filter(!pathogen_species %in% c("Multiple", "Puumala virus/Saaremaa virus")) %>%
  arrange(pathogen_family, pathogen_species, genus, species) %>%
  mutate(association = case_when(number_positive >= 1 ~ TRUE,
                                 TRUE ~ FALSE),
         detection_method = case_when(detection_method == "pcr" ~ "Viral",
                                      detection_method == "antibody" ~ "Antibody"))

scaled_n <- preProcess(as.data.frame(log10(hp_associations$number_tested)), method = "range")
hp_associations$alpha <- predict(scaled_n, as.data.frame(log10(hp_associations$number_tested)))[, 1]

host_pathogen_plot <- hp_associations %>%
  ggplot(aes(x = pathogen_species, y = species)) +
  geom_tile(aes(fill = association, colour = known_zoonosis, alpha = alpha,
                text = paste0("Pathogen species: ", pathogen_species, "<br>",
                              "Host species: ", species, "<br>",
                              "Number tested: ", number_tested, "<br>",
                              "Number positive: ", number_positive, "<br>",
                              "Known zoonosis: ", known_zoonosis))) +
  scale_fill_viridis_d() +
  scale_colour_manual(values = c("white", "black")) +
  facet_wrap(~ detection_method) +
  guides(alpha = "none",
         colour = "none") +
  labs(fill = "H-P association",
       x = element_blank(),
       y = element_blank()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggplotly(host_pathogen_plot, tooltip = "text")

```